---
title: "Temperature Germination Profile of *Celosia argentea*"
subtitle: "Exploratory Analysis Across 35 Accessions and Five Temperature Regimes"
author: "Haruna Aremu"
date: "04-03-2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi     = 150
)
```

---

## Background

This analysis explores the germination behaviour of 35 *Celosia argentea*
accessions (coded Tca-1 to Tca-35) grown under five temperature regimes
(15°C, 20°C, 25°C, 30°C, and 35°C). Seeds were sown in controlled-environment
chambers at Aarhus University, Flakkebjerg Research Facility (February–April 2026),
with germination counts recorded daily from Day 2 to Day 12 After Sowing (DAS).

Three germination metrics are computed and visualised:

- **Germination Percentage (GP)** — the proportion of seeds that germinated
- **Mean Germination Time (MGT)** — the average time (in days) to germination,
  weighted by daily germination counts
- **Germination Speed Index (GI)** — a speed-weighted sum of daily counts
  (higher values indicate faster germination)

> **Note on raw data:** The dataset is not included in this repository. It was
> collected as part of an ongoing, unpublished research collaboration at Aarhus
> University and cannot be shared publicly without institutional clearance.
> The script is fully reproducible once a compatible dataset is supplied —
> see the Data Import section for the expected structure.

---

## Setup

### Packages

```{r packages}
# pacman::p_load installs missing packages automatically before loading
pacman::p_load(tidyverse, readxl, tidytext, here)
```

### Global Theme

A single `theme_set()` call applies a consistent visual style to all figures,
avoiding repetition across individual plot objects.

```{r global-theme}
theme_set(
  theme_bw(base_size = 13) +
    theme(
      axis.title            = element_text(size = 16),
      axis.text             = element_text(size = 14),
      plot.caption          = element_text(hjust = 0, size = 14,
                                           face = "bold", colour = "black"),
      plot.caption.position = "plot",
      legend.text           = element_text(size = 12),
      panel.grid            = element_blank()
    )
)
```

### Colour Palette

A named vector maps each temperature level to a fixed colour. Passing this
to `scale_fill_manual()` and `scale_colour_manual()` ensures colours are
consistent across all eight figures.

```{r colour-palette}
temp_colours <- c(
  "15" = "#818cf8",   # cool violet
  "20" = "#60a5fa",   # blue
  "25" = "#4ade80",   # green
  "30" = "#f59e0b",   # amber
  "35" = "#f87171",   # coral
  "40" = "#b91c1c"    # deep red
)
```

### Save Helper

A small wrapper around `ggsave()` reduces repetition and ensures all output
files land in the `outputs/` folder with consistent resolution and units.

```{r save-helper}
save_plot <- function(filename, plot, h = 7, w = 12, format = "png") {
  ggsave(
    paste0(filename, ".", format), plot,
    path   = here("outputs"),
    height = h, width = w,
    units  = "in", dpi = 300,
    bg     = "white"
  )
}
```

---

## Data Import and Transformation

### Import

The dataset is expected as an Excel file with one row per accession-temperature
combination and daily germination counts in columns named `DAS2`, `DAS3`, ...,
`DAS12`. Temperature 45°C is excluded from all analyses (insufficient
germination at that extreme).

```{r import}
# Replace the path below with your own file location
celo <- read_xlsx(here("data/celosia_temperature_stress_data.xlsx"),
                  range = "A1:M246")
```

### Reshape to Long Format

`pivot_longer()` converts the wide DAS columns into a tidy long format,
with one row per accession-temperature-day observation. `str_remove()`
strips the "DAS" prefix so `das` is numeric and usable on plot axes.

```{r reshape}
celo_new <- celo %>%
  mutate(
    Acc  = factor(Acc, levels = paste0("Tca- ", 1:35)),
    Temp = factor(Temp)
  ) %>%
  pivot_longer(
    cols      = matches("DAS"),
    names_to  = "das",
    values_to = "count"
  ) %>%
  mutate(das = as.numeric(str_remove(das, "DAS"))) %>%
  filter(Temp != 45)
```

### Derived Datasets

Three summary datasets are computed once here and reused across all figures,
avoiding redundant computation inside individual plot chunks.

```{r derived-datasets}
# ── Germination Percentage (GP) ───────────────────────────────
# Maximum cumulative count divided by total seeds sown (25 per pot)
germination_perc <- celo_new %>%
  group_by(Acc, Temp) %>%
  reframe(
    max_count = max(count, na.rm = TRUE),
    gp        = (max_count / 25) * 100
  )

# ── Mean Germination Time (MGT) ───────────────────────────────
# Weighted mean of DAS, weighted by daily germination increments.
# pmax() floors negative increments (data entry artefacts) to zero.
# NA is returned for accession-temperature combinations with zero
# total germination, avoiding division by zero.
mgt <- celo_new %>%
  arrange(Acc, Temp, das) %>%
  group_by(Acc, Temp) %>%
  mutate(daily = pmax(count - lag(count, default = 0), 0)) %>%
  reframe(
    total_germination = sum(daily, na.rm = TRUE),
    mgt = if_else(
      total_germination > 0,
      sum((daily * das) / total_germination, na.rm = TRUE),
      NA_real_
    )
  )

# ── Germination Speed Index (GI) ─────────────────────────────
# Sum of daily germination counts divided by the DAS at which they
# occurred. Higher values indicate faster, earlier germination.
gi <- celo_new %>%
  arrange(Acc, Temp, das) %>%
  group_by(Acc, Temp) %>%
  mutate(daily = pmax(count - lag(count, default = 0), 0)) %>%
  reframe(gi = sum(daily / das, na.rm = TRUE))
```

---

## Figures

### Figure 1 — Germination Percentage Heatmap

The heatmap gives the most complete overview: every accession at every
temperature in a single panel. Diverging fill (dark → blue → green) makes
the 50% midpoint immediately readable. White text is overlaid for precise
values.

```{r fig1-heatmap, fig.height=9, fig.width=12}
fig1_heatmap <- ggplot(germination_perc,
                       aes(Temp, fct_rev(Acc), fill = gp)) +
  geom_tile(colour = "black", linewidth = 0.3) +
  geom_text(aes(label = round(gp, 1)), size = 4.5, colour = "white") +
  scale_fill_gradient2(
    low      = "#1a2332",
    mid      = "#2563eb",
    high     = "#4ade80",
    midpoint = 50,
    limits   = c(0, 100),
    name     = "Germination (%)"
  ) +
  labs(
    x       = "Temperature (\u00b0C)",
    y       = "Accession",
    caption = "Figure 1: Germination Percentage (%) of 35 Celosia argentea Accessions Across Five Temperatures"
  ) +
  theme(panel.border = element_blank())

fig1_heatmap

```

---

### Figure 2 — GP Distribution by Temperature (Boxplot)

Moving from individual accessions to a population summary, this boxplot
shows how GP varies across accessions at each temperature. Outliers are
suppressed to keep the focus on the bulk distribution.

```{r fig2-gp-boxplot, fig.height=7, fig.width=12}
fig2_gp_boxplot <- ggplot(germination_perc,
                          aes(x = Temp, y = gp, fill = Temp)) +
  geom_boxplot(outliers = FALSE, show.legend = FALSE, width = 0.8) +
  scale_fill_manual(values = temp_colours) +
  labs(
    x       = "Temperature (°C)",
    y       = "Germination Percentage (%)",
    caption = "Figure 2: Distribution of Germination Percentage (%) Across Accessions by Temperature"
  )

fig2_gp_boxplot
save_plot("fig2_gp_boxplot", fig2_gp_boxplot)
```

---

### Figure 3 — Mean Daily Germination by Temperature Over Time

Grouped bar charts show when germination peaks at each temperature.
`position = "dodge"` separates temperatures within each DAS. Error bars
are ±1 SE computed over all accessions. This bridges the GP summary (Fig 2)
into the time-course dynamics shown in Fig 4.

```{r fig3-daily-germination, fig.height=7, fig.width=12}
fig3_daily_germination <- celo_new %>%
  arrange(Acc, Temp, das) %>%
  group_by(Acc, Temp) %>%
  mutate(daily = pmax(count - lag(count, default = 0), 0)) %>%
  group_by(Temp, das) %>%
  reframe(
    mean_daily = mean(daily, na.rm = TRUE),
    # NA-aware denominator: only non-missing observations contribute to SE
    se_daily   = sd(daily, na.rm = TRUE) / sqrt(sum(!is.na(daily)))
  ) %>%
  ggplot(aes(x = das, y = mean_daily, fill = Temp)) +
  geom_col(position = "dodge", colour = "black", linewidth = 0.2) +
  geom_errorbar(
    aes(ymin = mean_daily - se_daily, ymax = mean_daily + se_daily),
    position = position_dodge(0.9), width = 0.3
  ) +
  scale_fill_manual(values = temp_colours, name = "Temperature (°C)") +
  scale_x_continuous(breaks = seq(2, 12, 1)) +
  labs(
    x       = "Day After Sowing (DAS)",
    y       = "Mean Daily Germination (seeds)",
    caption = "Figure 3: Mean Daily Germination Count Across Accessions by Temperature"
  )

fig3_daily_germination
save_plot("fig3_daily_germination", fig3_daily_germination)
```

---

### Figure 4 — Cumulative Germination Curves per Accession

Faceted by accession (7 columns × 5 rows), each panel shows the full
time course of cumulative germination across temperatures. This is the
most data-dense figure and is best read alongside Fig 1 to connect
final GP values to the trajectories that produced them.

```{r fig4-trend, fig.height=12, fig.width=14}
fig4_trend <- ggplot(celo_new, aes(das, count, colour = Temp)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_point(size = 2, na.rm = TRUE) +
  facet_wrap(~Acc, ncol = 7) +
  scale_x_continuous(breaks = seq(2, 12, 2)) +
  scale_colour_manual(values = temp_colours, name = "Temperature (°C)") +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "grey92"),
    axis.title.x     = element_text(size = 12),
    axis.title.y     = element_text(size = 12),
    axis.text.x      = element_text(size = 9)
  ) +
  labs(
    x       = "Days After Sowing (DAS)",
    y       = "Cumulative Germinated Seeds",
    caption = "Figure 4: Cumulative Germination Curves for 35 Celosia argentea Accessions Across Five Temperatures"
  )

fig4_trend
save_plot("fig4_trend", fig4_trend, h = 12, w = 14, format = "tiff")
```

---

### Figure 5 — MGT per Accession (Faceted Bar Chart)

Shifting from *how many* germinated to *how fast*, this figure mirrors
the structure of Fig 1 at the accession level. Shorter bars indicate
faster germination. Temperature 40°C is absent where germination was
zero (MGT returns NA and is silently dropped by `geom_col()`).

```{r fig5-mgt-accession, fig.height=12, fig.width=14}
fig5_mgt_accession <- ggplot(mgt, aes(Temp, mgt, fill = Temp)) +
  geom_col(na.rm = TRUE) +
  scale_fill_manual(values = temp_colours, name = "Temperature (°C)") +
  facet_wrap(~Acc, ncol = 7, scales = "free_x") +
  theme(
    legend.position  = "bottom",
    axis.text.y.left = element_text(size = 11.5)
  ) +
  labs(
    x       = "Temperature (°C)",
    y       = "Mean Germination Time (Days)",
    caption = "Figure 5: Mean Germination Time (MGT) of 35 Celosia argentea Accessions"
  )

fig5_mgt_accession
save_plot("fig5_mgt_accession", fig5_mgt_accession, h = 12, w = 14)
```

---

### Figure 6 — MGT Distribution by Temperature (Violin + Boxplot)

The violin layer reveals the full distribution shape across accessions;
the inner boxplot provides median and IQR. Together they show both the
central tendency and spread of germination speed at each temperature,
summarising Fig 5 in a single panel.

```{r fig6-mgt-violin, fig.height=7, fig.width=12}
fig6_mgt_violin <- ggplot(mgt, aes(Temp, mgt, fill = Temp)) +
  geom_violin(width = 0.8, alpha = 0.4, show.legend = FALSE) +
  geom_boxplot(width = 0.2, show.legend = FALSE, outliers = FALSE) +
  scale_fill_manual(values = temp_colours) +
  labs(
    x       = "Temperature (°C)",
    y       = "Mean Germination Time (Days)",
    caption = "Figure 6: Distribution of Mean Germination Time (MGT) Across Accessions by Temperature"
  )

fig6_mgt_violin
save_plot("fig6_mgt_violin", fig6_mgt_violin)
```

---

### Figure 7 — GI per Accession (Faceted Bar Chart)

The Germination Speed Index complements MGT: while MGT measures *when*
germination occurs, GI weights seeds that germinate earlier more heavily,
giving a single score that captures both speed and volume. This figure
mirrors the structure of Fig 5 for direct comparison.

```{r fig7-gi-accession, fig.height=12, fig.width=14}
fig7_gi_accession <- ggplot(gi, aes(Temp, gi, fill = Temp)) +
  geom_col() +
  scale_fill_manual(values = temp_colours, name = "Temperature (°C)") +
  facet_wrap(~Acc, ncol = 7) +
  theme(legend.position = "bottom") +
  labs(
    x       = "Temperature (°C)",
    y       = "Germination Speed Index",
    caption = "Figure 7: Germination Speed Index (GI) of 35 Celosia argentea Accessions Across Five Temperatures"
  )

fig7_gi_accession
save_plot("fig7_gi_accession", fig7_gi_accession, h = 12, w = 14, format = "tiff")
```

---

### Figure 8 — GI Distribution by Temperature (Boxplot)

The closing figure summarises GI across accessions per temperature,
completing the GP → MGT → GI narrative arc. Higher GI and lower MGT
at the same temperature confirm faster, more vigorous germination —
the two metrics should be interpreted together.

```{r fig8-gi-boxplot, fig.height=7, fig.width=12}
fig8_gi_boxplot <- ggplot(gi, aes(Temp, gi, fill = Temp)) +
  geom_boxplot(width = 0.8, outliers = FALSE, show.legend = FALSE) +
  scale_fill_manual(values = temp_colours) +
  labs(
    x       = "Temperature (°C)",
    y       = "Germination Speed Index",
    caption = "Figure 8: Distribution of Germination Speed Index (GI) Across Accessions by Temperature"
  )

fig8_gi_boxplot
save_plot("fig8_gi_boxplot", fig8_gi_boxplot)
```

---

## Summary

```{r summary-table, echo=FALSE}
germination_perc %>%
  group_by(Temp) %>%
  reframe(
    Mean_GP = round(mean(gp, na.rm = TRUE), 1),
    SD_GP   = round(sd(gp, na.rm = TRUE), 1),
    Min_GP  = round(min(gp, na.rm = TRUE), 1),
    Max_GP  = round(max(gp, na.rm = TRUE), 1)
  ) %>%
  knitr::kable(
    col.names = c("Temperature (°C)", "Mean GP (%)", "SD", "Min (%)", "Max (%)"),
    caption   = "Table 1: Summary of Germination Percentage by Temperature"
  )
```

**Key findings:**

- Optimal temperature for germination capacity (GP): **35°C**
- Optimal temperature for germination speed (MGT / GI): **30°C**
- Germination at 15°C was consistently low across all accessions
- Considerable inter-accession variation is visible at all temperatures,
  suggesting heritable differences in thermal tolerance worth further investigation


