---
title: "week3-2 Assignment - Chi-Square Test"
subtitle: "Haruna Aremu"
date: "06-04-2026"
output: html_document
---

This analysis applies chi-square goodness-of-fit tests to evaluate whether molecular 
markers in two populations — mouse and maize — conform to their expected Mendelian 
segregation ratios. The mouse dataset contains marker genotype data for progeny derived 
from a backcross population, where each marker is expected to segregate in a 1:1 ratio. 
The maize dataset contains marker genotype data for an F2 population derived from 
divergent parents, where markers are expected to follow a 1:2:1 ratio. Markers that 
significantly deviate from these expectations are considered to show segregation distortion.

# Loading of Data

The datasets are loaded from CSV files and previewed using the `head()` function to 
confirm their structure before analysis.

```{r mouse data}
mouse_data <- read.csv("data/mouse.csv")

head(mouse_data)

```


```{r maize data}
maize_data <- read.csv("data/maize.csv")

head(maize_data)
```
# Performing Chi-square Test

## Mouse Data

For a backcross population, each marker is expected to segregate in a **1:1 ratio** 
between the two parental allele classes. A chi-square goodness-of-fit test is applied 
to each marker to determine whether the observed genotype counts deviate significantly 
from this expectation. P-values are stored for all 14 markers and plotted in ascending 
order against the conventional significance threshold of α = 0.05.

```{r}
# For mouse data, the expected ratio is 1:1

exp <- c(0.5, 0.5)
p_value <- c()

M <- ncol(mouse_data[, 2: 15])

for (m in 2: 15) {
  obs <- table(mouse_data[ , m])
  
  chi_test <- chisq.test(x = obs, p = exp)
  
  p_value[m -1] <- chi_test$p.value
  
}


```

The sorted p-values are plotted below. Markers falling below the red line (α = 0.05) 
show statistically significant deviation from the expected 1:1 segregation ratio

```{r}

plot(sort(p_value))
abline(h = 0.05, col = "red")
legend("topleft", legend=c("alpha = 0.05"), col=c("red"), lty=1)
```

A frequency table is used to count how many markers fall below the significance 
threshold, indicating segregation distortion.

```{r}
table(p_value < 0.05)
```
Only 1 marker shows significant deviation from the expected 1:1 segregation ratio at 
α = 0.05, suggesting that the vast majority of markers in this backcross population 
are segregating normally.

## Bonferroni Correction

When testing multiple markers simultaneously, the probability of obtaining at least one 
false positive increases with the number of tests performed. The Bonferroni correction 
addresses this multiple testing problem by adjusting the significance threshold to 
α/M, where M is the total number of markers tested. This is a conservative approach 
that controls the family-wise error rate.

```{r}
table(p_value < 0.05/M)
```

The plot below overlays both the uncorrected (α = 0.05, red) and Bonferroni-corrected 
(α/M, blue) thresholds, allowing a visual comparison of which markers remain significant 
after the correction is applied.

```{r}
plot(sort(p_value))
abline(h = c(0.05, 0.05/M), col = c("red", "blue"))
legend("topleft", legend=c("alpha = 0.05"), col=c("red"), lty=1)
```


## Maize Data

For an F2 population, heterozygous parents are expected to produce offspring in a 
**1:2:1 ratio** (homozygous reference : heterozygous : homozygous alternate), consistent 
with Mendelian segregation at a single locus. Chi-square goodness-of-fit tests are 
applied across all 12 markers, and p-values are plotted against the α = 0.05 threshold.

```{r}
exp_maize <- c(0.25, 0.5, 0.25)
p_value_maize <- c()
M_maize <- ncol(maize_data[, 2:13])

for(m in 2:13){
  obs <- table(maize_data[, m])
  chi_test <- chisq.test(x = obs, p = exp_maize)
  p_value_maize[m - 1] <- chi_test$p.value
}

plot(sort(p_value_maize))
abline(h = 0.05, col = "red")
```

A frequency table is used to count how many markers fall below the significance 
threshold, indicating segregation distortion

```{r}
table(p_value_maize < 0.05)
```

Here all  12 markers in the maize F2 population have p-values exceeding the 0.05 significance 
threshold, indicating that none deviate significantly from the expected 1:2:1 segregation ratio. Since no evidence of segregation distortion is detected, applying a Bonferroni 
correction is unnecessary in this case.



















































































